\textbf{Unshuffling Permutations}

### 2016-04-06 slides/slides.tex
Suite

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Combinatorial products and structure coefficients}
Let $C$ be a set of combinatorial objects and let $\QQ[C]$ be the
$\QQ$-linear span of $C$.
\medskip

@@

In a general way, for any $x, y \in C$,
\begin{equation*}
    x \cdot y = \sum_{z \in C} \lambda_{x, y}^z \; z.
\end{equation*}

A product is wholly encoded by the coefficients
$\lambda_{x, y}^z \in \QQ$, $x, y, z \in C$, called
\alert{structure coefficients}.
\end{frame}

@@

In a general way, for any $z \in C$,
\begin{equation*}
    \Delta(x) = \sum_{x, y \in C}  \lambda_{x, y}^z \;  x \otimes y.
\end{equation*}

A coproduct is wholly encoded by the coefficients
$\lambda_{x, y}^z \in \QQ$, $x, y, z \in C$, called
\alert{structure coefficients}.
\end{frame}

@@

Conversely, if $\Delta : \QQ[C] \to \QQ[C] \otimes \QQ[C]$ is
a coproduct with structure coefficients $\lambda_{x, y}^z \in \QQ$,
$x, y, z \in C$, we can define its \alert{dual product} $\cdot$ by
setting
\begin{equation*}
    x \cdot y := \sum_{z \in C} \lambda_{x, y}^z \; z.
\end{equation*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%o%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Example: shuffling binary words}
\begin{block}{Proposition}
    The shuffle product of binary words is the dual product of the
    unshuffling coproduct of binary words.

        REF ? SCHUTZ ?
\end{block}
\medskip

\begin{block}{Example}
\begin{equation*}
    \Delta(100) =
    \epsilon \otimes 100 + 1 \otimes 00
    + \textcolor{BrickRed}{2 (0 \otimes 10)} + 2 (10 \otimes 0) + 00 \otimes 1
    + 100 \otimes \epsilon
\end{equation*}
\smallskip

Since $0 \otimes 10$ has multiplicity $2$ in $\Delta(100)$, by duality,
$100$ has multiplicity $2$ in $0 \shuffle 10$:
\begin{equation*}
    0 \shuffle 10 = 010 + \textcolor{BrickRed}{2 (100)}.
\end{equation*}
\end{block}
\end{frame}

@@

This shuffle product has been introduced for the study of some
Hopf algebra \BIB{Vargas}{2014}.
\medskip

It is a kind of association of the convolution product and the shifted
shuffle product of permutations \BIB{Duchamp, Hivert, Thibon}{2002}.
\end{frame}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Combinatorial products and structure coefficients}
Let $C$ be a set of combinatorial objects and $\QQ[C]$ be the
$\QQ$-linear span of $C$.
\medskip

@@

In a general way, for any $x, y \in C$,
\begin{equation*}
    x \cdot y = \sum_{z \in C} \lambda_{x, y}^z \; z.
\end{equation*}
\medskip

A product is therefore wholly encoded by the coefficients
$\lambda_{x, y}^z \in \QQ$, $x, y, z \in C$, called
\alert{structure coefficients}.
\end{frame}

@@

In a general way, for any $z \in C$,
\begin{equation*}
    \Delta(x) = \sum_{x, y \in C}  \lambda_{x, y}^z \;  x \otimes y.
\end{equation*}
\medskip

A coproduct is therefore wholly encoded by the coefficients
$\lambda_{x, y}^z \in \QQ$, $x, y, z \in C$, called
\alert{structure coefficients}.
\end{frame}

@@

Conversely, if $\Delta : \QQ[C] \to \QQ[C] \otimes \QQ[C]$ is
a coproduct with structure coefficients $\lambda_{x, y}^z \in \QQ$,
$x, y, z \in C$, we can define its \alert{dual product} $\cdot$ by
setting
\begin{equation*}
    x \cdot y := \sum_{z \in C} \lambda_{x, y}^z \; z.
\end{equation*}
\medskip

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%o%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Example: shuffling binary words}
\begin{block}{Proposition \BIB{Reutenauer}{1993}}
    The shuffle product of binary words is the dual product of the
    unshuffling coproduct of binary words.
\end{block}
\medskip

\begin{block}{Example}
Since
\begin{equation*}
    \textcolor{RoyalBlue}{\Delta(100)} =
    \epsilon \otimes 100 + 1 \otimes 00
    + \textcolor{Orange}{2} (\textcolor{BrickRed}{0 \otimes 10})
    + 2 (10 \otimes 0) + 00 \otimes 1
    + 100 \otimes \epsilon,
\end{equation*}
we observe that $\textcolor{BrickRed}{0 \otimes 10} $ has multiplicity
$\textcolor{Orange}{2}$ in $\textcolor{RoyalBlue}{\Delta(100)}$.
\medskip

By duality, this implies that $\textcolor{RoyalBlue}{100}$ has
multiplicity $\textcolor{Orange}{2}$ in
$\textcolor{BrickRed}{0 \shuffle 10}$. Indeed,
\begin{equation*}
    \textcolor{BrickRed}{0 \shuffle 10} =
    010 + \textcolor{Orange}{2} (\textcolor{RoyalBlue}{100}).
\end{equation*}
\end{block}
\end{frame}

@@

Product first introduced for the study of some Hopf algebras~\BIB{Vargas}{2014}.
\medskip

Similar to the convolution product and the shifted
shuffle product \BIB{Duchamp, Hivert, Thibon}{2002}.
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Square permutations}
\begin{block}{Definition}
    A permutation $\pi$ is a \alert{square} (w.r.t. the shuffle product
    $\SHUFFLE$) if there is $\sigma \in S$ such that $\pi$ appears in
    $\sigma \SHUFFLE \sigma$.
\end{block}
\medskip

By duality, $\pi$ is a square if there is $\sigma \in S$ such that
$\sigma \otimes \sigma$ appears in $\Delta(\pi)$.
\medskip

\begin{block}{Proposition}
    A permutation $\pi$ is a square iff $\pi$ can be obtained by
    shuffling two order-isomorphic words.
\end{block}
\medskip

\begin{block}{Example}
    The permutation
    \begin{math}
        \pi :=
        \textcolor{RoyalBlue}{25}\textcolor{BrickRed}{16}
        \textcolor{RoyalBlue}{7}\textcolor{BrickRed}{83}
        \textcolor{RoyalBlue}{4}
    \end{math}
    is square since $\pi$ can be obtained by shuffling
    $\textcolor{BrickRed}{1683}$ and $\textcolor{RoyalBlue}{2574}$.
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Some properties of square permutations}
The first numbers of square permutations are, size by size,
\begin{equation*}
    1, 0, 2, 0, 20, 0, 504, 0, 21032, 0, 1293418.
\end{equation*}
\medskip

\begin{block}{Proposition}
    Let $\pi$ be a square permutation and $\sigma$ be a square root of
    $\pi$. Then,
    \begin{enumerate}
        \item $\widetilde{\pi}$ is a square and
        $\widetilde{\sigma}$ is one of its square roots;
        \item $\bar \pi$ is a square and $\bar \sigma$ is one of
        its square roots;
        \item $\pi^{-1}$ is a square and $\sigma^{-1}$ is one of
        its square roots.
    \end{enumerate}
\end{block}
\medskip

\begin{block}{Proposition}
    The set of binary words of length $n$ that are square w.r.t.
    $\shuffle$ is in one-to-one correspondence with the set of
    square permutations of length $n$ avoiding the patterns
    $213$ and~$231$.
\end{block}
